Density-dependent selection and the limits of relative fitness

### 2018-02-23 manuscript.tex
l

The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'') i.e. $x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules. The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$. Thus 
\begin{equation}
\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
\end{equation}
where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$ \textit{after} dispersal (the significance of evaluating after dispersal should become clearer below), in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.

Our approximation is to replace $x_j$ with its mean in the last term in Eq.~\eqref{eq:deltr},
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}}.\label{eq:meanfieldr}
\end{equation}
Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.

@@

The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal. $\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence). As a result, if the focal type is rare, $\tilde{q}$ predicts that the mean propagule density is $L+1$ in territories with one focal propagule. This is erroneous, because the mean number of propagules per territory when lottery competition commences is $L$, regardless of whether a focal propagule happens to be present. The discrepancy causes rare invaders evaluated under $\tilde{q}$ to have an intrinsic rarity disadvantage, which is absurd (i.e. even if $b$, $c$ and $d$ are identical, $\tilde{q}$ predicts that a rare type will decline in frequency). By contrast, Eq. \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''. By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations. The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.

>>>>

The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'') i.e. $x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules. The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$. Thus 
\begin{equation}
\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
\end{equation}
where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.

To evaluate the expectation in Eq. \eqref{eq:deltr}, we replace the $x_j$ with  ``effective'' mean values $\langle x_j\rangle_{\tilde{q}}$, to obtain the approximation
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{q}}}.\label{eq:meanfieldr}
\end{equation}
Note that the effective means $\langle x_j\rangle_{\tilde{q}}$ are taken with respect to a new distribution $\tilde{q}$. We will define $\tilde{q}$ below; first we explain the motivation for introducing it. Suppose that we use $\tilde{p}$ instead of $\tilde{q}$ in Eq. \eqref{eq:meanfieldr} ($x_j\rightarrow \langle x_j\rangle_{\tilde{p}}$). Moreover, suppose that the focal type is rare ($l_i\ll 1$), that the total propagule density is high $L\gg 1$, and that all types have identical $b$, $c$ and $d$. Then the mean total propagule density in territories with one focal propagule is $1+\sum_j \langle x_j\rangle_{\tilde{p}}\approx L+1$ and the right hand side of Eq. \eqref{eq:meanfieldr} is approximately $1/(L+1)$. Since $L$ is large and the focal type is rare, a negligible fraction of territories will be have a single uncontested focal propagule or more than one focal propagule. Thus, $1/(L+1)$ is the probability that a random focal propagule successfully wins a territory in the Eq. \eqref{eq:meanfieldr} approximation. On the other hand, virtually all nonfocal propagules

 territories

\label{eq:meanfieldr}

Since $x_i$ and the $x_j$ are independent under $\tilde{p}$

If the focal type is rare, the mean propagule density is $L+1$ in territories with one focal propagule.

The discrepancy causes rare invaders evaluated under $\tilde{q}$ to have an intrinsic rarity disadvantage, which is absurd (i.e. even if $b$, $c$ and $d$ are identical, $\tilde{q}$ predicts that a rare type will decline in frequency).

By contrast, Eq. \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''. By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations. The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.

$\tilde{p}$

The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal. $\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).

Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.

@@
